---
title: "Creating Copy-Paste Friendly Table Summaries in R"
author: "Raymond Balise"
date: last-modified
format: html
vignette: >
  %\VignetteIndexEntry{Creating Copy-Paste Friendly Table Summaries in R}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
knitr:
  opts_chunk:        ########## set global options ############
    collapse: true   # keep code from blocks together (if shown)
    echo: false      # don't show code
    message: true    # show messages
    warning: true    # show warnings
    error: true      # show error messages
    comment: ""      # don't show ## with printed output
    dpi: 100         # image resolution (typically 300 for publication)
    fig-width: 6.5   # figure width
    fig-height: 4.0  # figure height
    R.options:    
      digits: 3      # round to three digits
editor: source
---

```{r}
#| echo: false

library(tibble)
library(hms)

# Create a tibble with all common R data types for testing
test_data <- tibble(
  # Character
  char_col = c("apple", "banana", "cherry", "damson", "elderberry"),

  # Numeric (double)
  numeric_col = c(1.5, 2.7, 3.14, 4.0, 5.9),
  numeric_col_miss = c(1.5, 2.7, 3.14, NA, 5.9),

  # Integer
  integer_col = c(1L, 2L, 3L, 4L, 5L),
  integer_col_miss = c(1L, 2L, 3L, NA, 5L),

  # Logical
  logical_col = c(TRUE, FALSE, TRUE, FALSE, TRUE),
  logical_col_miss = c(TRUE, FALSE, TRUE, NA, TRUE),

  # Factor (unordered)
  factor_col = factor(c("low", "medium", "high", "medium", "low")),
  factor_col_miss = factor(c("low", "medium", "high", NA, "low")),

  # Ordered factor
  ordered_col = ordered(c("small", "medium", "large", "medium", "small"),
                        levels = c("small", "medium", "large")),
  # Ordered factor
  ordered_col_miss = ordered(c("small", "medium", "large", NA, "small"),
                        levels = c("small", "medium", "large")),

  # Date
  date_col = as.Date(c("2034-01-01", "2034-06-15", "2034-12-31", "2033-03-20", "2035-08-10")),

  # POSIXct (datetime)
  datetime_col = as.POSIXct(c("2034-01-01 09:30:00", "2034-06-15 14:45:30",
                              "2034-12-31 23:59:59", "2033-03-20 08:15:22",
                              "2035-08-10 16:20:45")),

  # hms (time only)
  time_col = hms::as_hms(c("09:30:00", "14:45:30", "23:59:59", "08:15:22", "16:20:45")),
  time_col_miss = hms::as_hms(c("09:30:00", "14:45:30", "23:59:59", NA, "16:20:45")),

  # Complex numbers
  # complex_col = c(1+2i, 3+4i, 5+0i, 0+6i, 2-3i),

  # Raw bytes
  # raw_col = as.raw(c(65, 66, 67, 68, 69)),  # A, B, C, D, E in ASCII

  # List column (nested data)
  # list_col = list(
  #   c(1, 2, 3),
  #   c("a", "b"),
  #   c(TRUE, FALSE, TRUE),
  #   NULL,
  #   data.frame(x = 1:2, y = letters[1:2])
  # )
)
```

# The Problem


To get the most helpful answers to analysis questions from AI tools (like ChatGPT, Claude, Copilot, and Gemini), the tools need to know about your data but uploading datasets is often problematic or even illegal. Other summaries like the `skim()` function in the [skimr](https://docs.ropensci.org/skimr/) package or `str()` from base R will print potentially sensitive data like character strings or dates.  To learn more about de-identification and protected health information (PHI) in the United States, visit the Health and Human Services webpage on [De-Identification of PHI](https://www.hhs.gov/hipaa/for-professionals/special-topics/de-identification/index.html#rationale).

# The Solution

It would be useful to have a function that prints a description of datasets that **excludes** details that are known to be, or are likely to be, sensitive.  For example, dates are protected health information; free-form text is also problematic.  While the output from the function needs to be checked to make sure there is no sensitive data, it is useful to have a function that prints variable names, variable types and the values for categorical data.

To have R print a description of your dataset — for example, for a dataset named `test_data` — you can first tell R to load the `Open.Visualization.Academy` package into its thinking memory and then use the `show_structure()` function like this:
```{.r}
library(Open.Visualization.Academy)
show_structure(test_data)
```

or you can tell R to use the function with only one line, like this:

```{.r}
Open.Visualization.Academy::show_structure(test_data)
```

The result will look like this:

```{.markdown}
Table: `test_data` looks like this

|variable         |type               |levels                           |
|:----------------|:------------------|:--------------------------------|
|char_col         |character          |                                 |
|numeric_col      |numeric            |range: [1.5-5.9]                 |
|numeric_col_miss |numeric            |range: [1.5-5.9], NA             |
|integer_col      |integer            |range: [1-5]                     |
|integer_col_miss |integer            |range: [1-5], NA                 |
|logical_col      |logical            |TRUE, FALSE                      |
|logical_col_miss |logical            |TRUE, FALSE, NA                  |
|factor_col       |factor             |high, low, medium                |
|factor_col_miss  |factor             |high, low, medium, NA            |
|ordered_col      |ordered factor     |small, medium, large             |
|ordered_col_miss |ordered factor     |small, medium, large, NA         |
|date_col         |Date               |                                 |
|datetime_col     |datetime           |                                 |
|time_col         |time (hrs:min:sec) |range: [08:15:22 - 23:59:59]     |
|time_col_miss    |time (hrs:min:sec) |range: [09:30:00 - 23:59:59], NA |
✔ Copied to the clipboard!
Remove any sensitive data before pasting and sharing.
Look for: names, dates, locations, phone numbers, IDs, emails, etc.

! Review factor levels for sensitive information:
factor_col, factor_col_miss, ordered_col, ordered_col_miss
```

If your operating system allows you to copy and paste, the report will be copied automatically onto your clipboard. 

The report is designed to not print sensitive data like names (which are likely character variables) and dates. It will print the names of categorical factor variables along with their levels. The bottom of the report lists categorical factor variables which contain text *other* than: `"yes", "no", "checked", "unchecked", "TRUE", "FALSE", "male", "female"`. Carefully check these variables for potentially sensitive information before pasting the output into any AI tools or sharing with the public.

## Extra option: `display_redacted`

By default, `show_structure()` hides/suppresses/redacts character variables, dates, and datetime variables from the levels column.  We prefer this format because it makes it quicker to review the output. However, if you don't like to see blank lines for the redacted data, use the `display_redacted = TRUE` argument within the `show_structure()` function. Setting the value to `TRUE` prints `< redacted strings >` for variables of type "character", `< redacted dates >` for date variables, and `< redacted date-times >` for datetime variables.

```{r}
#| echo: false
example_data <- tibble::tibble(
  mrn = rep(123456789) + 1:7,
  sex = factor(c("Male", "Male", "Female", "Male", "Male", "Male", "Male")),
  first_name = c("Kyle", "Raymond", "Lori", "Danny", "Dan", "Sean", "Gabriel"),
  last_name = factor(
    c("Grealis", "Balise", "Balise", "Maya", "Feaster", "Luo", "Odom")
  ),
  city = factor(
    c("Dallas", "Miami", "Miami", "Coral Gables", "Dallas", "New York", "Miami")
  ),
  package_author = factor(
    c("this", "this", "none", "none", "none", "none", "other")
  ),
  visit_date = as.Date(
    c(
      "2034-01-15", "2034-02-20", "2034-02-20", "2034-03-10", "2034-04-05", 
      "2034-05-12", "2034-06-18"
    )
  )
)
```

For example:

* `display_redacted = TRUE`:

```{.markdown}
Table: `example_data` looks like this

|variable       |type      |levels                                    |
|:--------------|:---------|:-----------------------------------------|
|mrn            |numeric   |range: [123456790-123456796]              |
|sex            |factor    |Female, Male                              |
|first_name     |character |< redacted strings >                      |
|last_name      |factor    |Balise, Feaster, Grealis, Luo, Maya, Odom |
|city           |factor    |Coral Gables, Dallas, Miami, New York     |
|package_author |factor    |none, other, this                         |
|visit_date     |Date      |< redacted dates >                        |
✔ Copied to the clipboard!
Remove any sensitive data before pasting and sharing.
Look for: names, dates, locations, phone numbers, IDs, emails, etc.

! Review factor levels for sensitive information:
last_name, city, package_author
```

* `display_redacted = FALSE` (default):

```{.markdown}
Table: `example_data` looks like this

|variable       |type      |levels                                    |
|:--------------|:---------|:-----------------------------------------|
|mrn            |numeric   |range: [123456790-123456796]              |
|sex            |factor    |Female, Male                              |
|first_name     |character |                                          |
|last_name      |factor    |Balise, Feaster, Grealis, Luo, Maya, Odom |
|city           |factor    |Coral Gables, Dallas, Miami, New York     |
|package_author |factor    |none, other, this                         |
|visit_date     |Date      |                                          |
✔ Copied to the clipboard!
Remove any sensitive data before pasting and sharing.
Look for: names, dates, locations, phone numbers, IDs, emails, etc.

! Review factor levels for sensitive information:
last_name, city, package_author
```

Note that `show_structure()` is not smart enough to notice that some numeric values, like the medical record number variable named `mrn`, and factor levels for `last_name` and `city` may be sensitive.  Carefully check the report and remove **_all_** sensitive data before pasting and sharing.

```{r}
suppressPackageStartupMessages(library(dplyr))
suppressPackageStartupMessages(library(kableExtra))

options(scipen = 999)
example_data |> 
  kable(format = "html", escape = FALSE) |> 
  kable_styling(bootstrap_options = c("striped", "hover")) |> 
  column_spec(c(1, 4, 5), color = "red", include_thead = TRUE)
options(scipen = 0)

```

So, before sharing the report you would want to edit it to show this:

```{.markdown}
Table: `example_data` looks like this

|variable       |type      |levels                                    |
|:--------------|:---------|:-----------------------------------------|
|mrn            |numeric   |                                          |
|sex            |factor    |Female, Male                              |
|first_name     |character |                                          |
|last_name      |factor    |                                          |
|city           |factor    |                                          |
|package_author |factor    |none, other, this                         |
|visit_date     |Date      |                                          |
```

## What is the example data?

If you are curious, the `test_data` used for the first report above contains all the types of data you are likely to see. Notice there are columns that were designed to have no missing data (like `numeric_col`) and columns that contain missing values (like `numeric_col_miss`).

```{r}
# Function to replace NA with red-colored NA in HTML output
format_na_red <- function(x) {
  if_else(is.na(x), 
          '<span style="color: red;">NA</span>', 
          as.character(x))
}

# Apply the formatting and create the table
test_data |> 
  mutate(across(everything(), format_na_red)) |> 
  kable(format = "html", escape = FALSE) |> 
  kable_styling(bootstrap_options = c("striped", "hover"))
```

