---
title: "Creating Copy-Paste Friendly Table Summaries in R"
author: "Raymond Balies"
date: last-modified
format: html
vignette: >
  %\VignetteIndexEntry{Creating Copy-Paste Friendly Table Summaries in R}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
knitr:
  opts_chunk:        ########## set global options ############
    collapse: true   # keep code from blocks together (if shown)
    echo: false      # don't show code
    message: true    # show messages
    warning: true    # show warnings
    error: true      # show error messages
    comment: ""      # don't show ## with printed output
    dpi: 100         # image resolution (typically 300 for publication)
    fig-width: 6.5   # figure width
    fig-height: 4.0  # figure height
    R.options:    
      digits: 3      # round to three digits
editor: source
---

```{r}
#| echo: false

library(tibble)
library(hms)

# Create a tibble with all common R data types for testing
test_data <- tibble(
  # Character
  char_col = c("apple", "banana", "cherry", "damson", "elderberry"),

  # Numeric (double)
  numeric_col = c(1.5, 2.7, 3.14, 4.0, 5.9),
  numeric_col_miss = c(1.5, 2.7, 3.14, NA, 5.9),

  # Integer
  integer_col = c(1L, 2L, 3L, 4L, 5L),
  integer_col_miss = c(1L, 2L, 3L, NA, 5L),

  # Logical
  logical_col = c(TRUE, FALSE, TRUE, FALSE, TRUE),
  logical_col_miss = c(TRUE, FALSE, TRUE, NA, TRUE),

  # Factor (unordered)
  factor_col = factor(c("low", "medium", "high", "medium", "low")),
  factor_col_miss = factor(c("low", "medium", "high", NA, "low")),

  # Ordered factor
  ordered_col = ordered(c("small", "medium", "large", "medium", "small"),
                        levels = c("small", "medium", "large")),
  # Ordered factor
  ordered_col_miss = ordered(c("small", "medium", "large", NA, "small"),
                        levels = c("small", "medium", "large")),

  # Date
  date_col = as.Date(c("2034-01-01", "2034-06-15", "2034-12-31", "2033-03-20", "2035-08-10")),

  # POSIXct (datetime)
  datetime_col = as.POSIXct(c("2034-01-01 09:30:00", "2034-06-15 14:45:30",
                              "2034-12-31 23:59:59", "2033-03-20 08:15:22",
                              "2035-08-10 16:20:45")),

  # hms (time only)
  time_col = hms::as_hms(c("09:30:00", "14:45:30", "23:59:59", "08:15:22", "16:20:45")),
  time_col_miss = hms::as_hms(c("09:30:00", "14:45:30", "23:59:59", NA, "16:20:45")),

  # Complex numbers
  # complex_col = c(1+2i, 3+4i, 5+0i, 0+6i, 2-3i),

  # Raw bytes
  # raw_col = as.raw(c(65, 66, 67, 68, 69)),  # A, B, C, D, E in ASCII

  # List column (nested data)
  # list_col = list(
  #   c(1, 2, 3),
  #   c("a", "b"),
  #   c(TRUE, FALSE, TRUE),
  #   NULL,
  #   data.frame(x = 1:2, y = letters[1:2])
  # )
)
```

# The Problem

To get the most helpful answers to analysis questions from AI tools (like ChatGPT, Claude, Copilot, and Gemini), the tools need to know about your data but uploading datasets is often problematic or even illegal. To learn more about de-identification and protected health information (PHI) in the United States, visit the Health and Human Services webpage on [De-Identification of PHI](https://www.hhs.gov/hipaa/for-professionals/special-topics/de-identification/index.html#rationale).

# The Solution

It would be useful to have a function that prints a description of datasets that **excludes** details that are known to be, or are likely to be, sensitive.  For example, dates are protected health information, free-form text is also problematic.  While the output from the function needs to be checked to make sure there is no sensitive data, it is useful to have a function that prints variable names, variable types and the permitted values for categorical data.

To have R print a description of your dataset — for example, for a dataset named `test_data` — you can first tell R to load the `Open.Visualization.Academy` package into its thinking memory and then use the `show_structure()` function like this:
```{.r}
library(Open.Visualization.Academy)
show_structure(test_data)
```

or you can tell R to use the function with only one line, like this:

```{.r}
Open.Visualization.Academy::show_structure(test_data)
```

The result will look like this:

```{.markdown}
Table: `test_data` looks like this

|variable         |type               |levels                           |
|:----------------|:------------------|:--------------------------------|
|char_col         |character          |                                 |
|numeric_col      |numeric            |range: [1.5-5.9]                 |
|numeric_col_miss |numeric            |range: [1.5-5.9], NA             |
|integer_col      |integer            |range: [1-5]                     |
|integer_col_miss |integer            |range: [1-5], NA                 |
|logical_col      |logical            |TRUE, FALSE                      |
|logical_col_miss |logical            |TRUE, FALSE, NA                  |
|factor_col       |factor             |high, low, medium                |
|factor_col_miss  |factor             |high, low, medium, NA            |
|ordered_col      |ordered factor     |small, medium, large             |
|ordered_col_miss |ordered factor     |small, medium, large, NA         |
|date_col         |Date               |                                 |
|datetime_col     |datetime           |                                 |
|time_col         |time (hrs:min:sec) |range: [08:15:22 - 23:59:59]     |
|time_col_miss    |time (hrs:min:sec) |range: [09:30:00 - 23:59:59], NA |
✔ Copied to the clipboard!
Check for sensitive data before pasting and sharing.
! Look for names, IDs, and other sensitive information in the factor
  levels for: factor_col, factor_col_miss, ordered_col, ordered_col_miss
```

If your operating system allows you to copy and paste, the report will be copied automatically onto your clipboard. 

The report is designed to not print sensitive data like names (which are likely character variables) and dates. It will print the names of categorical factor variables along with their levels. The bottom of the report lists categorical factor variables which contain text *other* than: `"yes", "no", "checked", "unchecked", "TRUE", "FALSE", "male", "female"`. Carefully check these variables for potentially sensitive information before pasting the output into any AI tools or sharing with the public.

## What is the example data?

If you are curious, the example dataset has all the types of data which you are likely to see.  Notice there are columns which were made to have no missing data (like `numeric_col`) and columns that contain missing (like `numeric_col_miss`).

```{r}
suppressPackageStartupMessages(library(dplyr))
suppressPackageStartupMessages(library(kableExtra))

# Function to replace NA with red-colored NA in HTML output
format_na_red <- function(x) {
  if_else(is.na(x), 
          '<span style="color: red;">NA</span>', 
          as.character(x))
}

# Apply the formatting and create the table
test_data |> 
  mutate(across(everything(), format_na_red)) |> 
  kable(format = "html", escape = FALSE) |> 
  kable_styling(bootstrap_options = c("striped", "hover"))

# # Alternative approach using cell_spec from kableExtra
# test_data |> 
#   mutate(across(everything(), ~ cell_spec(.x, 
#                                           color = if_else(is.na(.x), "red", "black"),
#                                           format = "html"))) |> 
#   kable(format = "html", escape = FALSE) |> 
#   kable_styling(bootstrap_options = c("striped", "hover"))
```

## Extra option: `display_redacted`

Another option for the table display is to use the `display_redacted = TRUE` argument
within the `show_structure()` function. Setting the value to `TRUE` prints `< redacted
strings >` for variables of type "character" or `< redacted dates >` for date variables.
See this comparison:

* `display_redacted = FALSE` (default):

|variable    |type      |levels                                    |
|:-----------|:---------|:-----------------------------------------|
|id          |integer   |range: [1-6]                              |
|first_name  |character |                                          |
|some_factor |factor    |A, B, C, D, E, F                          |
|visit_date  |Date      |                                          |

* `display_redacted = TRUE`:

|variable    |type      |levels                                    |
|:-----------|:---------|:-----------------------------------------|
|id          |integer   |range: [1-6]                              |
|first_name  |character |< redacted strings >                      |
|some_factor |factor    |A, B, C, D, E, F                          |
|visit_date  |Date      |< redacted dates >                        |
